\thetitle

### 2008-06-14 paper.tex
Work on discussion.

% The model of flow behavior implicit in non-negative matrix factoriation technique is that flow behavior is a composite of some limited set of basic behaviors---and that fractions of each flow ``belong'' to each behavior. This is not an \latin{a priori} unreasonable hypothesis. Contrast that with the assumptions underlying uniform and marginal simplifications of traffic behavior, which can be seen to be intuitively unreasonable without ever conducting and actual experiment or calculation.

\keywords{
  traffic modeling \and
  traffic analysis \and
  workload generation \and
  wireless networks \and
  wireless simulation \and
  linear algebra \and
  non-negative matrix factorization
}

@@

Once traffic is represented linearly, alternatives to assuming uniformity or marginality become apparent.
We argue that matrix factorization, which is nondestructive and does not assume uniformity or marginality, is far superior to multiplication for this problem.
To demonstrate this approach, we use non-negative matrix factorization (\caps{NMF})~\cite{Lee01} to find a low-dimensional approximation of the packet size and inter-packet interval matrices of a real traffic trace.
This factorization extracts a small set of ``basic behaviors'' from the traffic matrix and simultaneously explains each flow's observed behavior as a mixture of these basic behaviors.
Although the factorization is computed without examining port numbers, basic behaviors nevertheless correspond to intuitive expectations for protocols.
We find basic behaviors corresponding to typical network activities, including file transfer, typing via \caps{SSH}, web surfing, ping traffic, peer-to-peer and others.

@@

Thus far, we have addressed two out of three of the originally posed questions regarding traffic modeling. First, we have proposed a general approach for representing whole-network traffic without making assumptions of uniformity or marginality. To this end, we introduced the concept of a linear representation of traffic. The general matrix model entails a specific linear representation, such that network traffic patterns can easily be expressed as matrices. Our experimental results show that the \caps{GMM} is sufficiently realistic for generating wireless workloads derived from traces. When uniform or marginal assumptions are applied, however, sufficient realism eludes us, indicating that if we are to simplify and reduce the representation of network behaviors, we must do some through some other technique. In \Section{modeling-concepts}, we observed that application of uniformity and marginality assumptions to traffic instances is equivalent, in the \caps{GMM}, to right and left multiplication, respectively, by matrices of rank one. These transformations destroy almost all of the information in the original traffic matrix---only the zero matrix has lower rank than these.

@@

The question that remains to be answered is how we can simplify and reduce the representation of network behaviors without making unwarranted and intuitively clearly false assumptions of uniformity of marginality of network behaviors. We present here preliminary results from applying non-negative matrix factorization to the packet behaviors of collections of flows. Recall from \Section{general-matrix-model}, that the matrices $Z$ and $V$ represent the packet size and inter-packet interval histograms of a collection of flows, with each row corresponding to a flow, and each column corresponding to a packet size or a range of inter-packet intervals. We apply \caps{NMF} to the concatenation of these matrices, $\mat{P}=[\mat{Z}~\mat{V}]$. \caps{NMF} attempts to approximate $\mat{P}$ as the product of three non-negative matrices:
\begin{align}
\mat{P} \approx \mat{W}\mat{\Sigma}\mat{H},
\end{align}
where $\mat{W} \in \R^{f \times K}$ is row-stochastic,
$\mat{\Sigma} \in \R^{K \times K}$ is diagonal with descending diagonal entries, and
$\mat{H} \in \R^{K \times (d_z+d_v)}$ is row-stochastic as well.
$K\in\N$ is the rank of this factorization, which must be determined somehow as well.
What is the interpretation of such factorization?
The rows of $\mat{H}$ are \emph{basic behaviors}---pairs of packet size and inter-packet interval distributions, such that each flow's packet behavior can be approximated as a linear combination of these basic behaviors.
Specifically, the matrix $\mat{M}\mat{\Sigma}$ gives the mixing coefficients for each flow.
The diagonal entries of $\mat{\Sigma}$ have a special meaning as well:
they are the number of packets associated with each basic behavior.
Thus the basic behaviors are sorted in descending order of how much traffic they explain.
Note that the factorization does not associate basic behaviors with individual packets;
rather, total flow behavior is explained as a mixture of basic behaviors, so the fraction of packets in each flow explained by a given behavior is known, while the specific packets are unknown.

The striking results of applying non-negative matrix factorization to $\mat{P}$ are shown in \Figure{basic-behaviors}. Twelve basic behaviors suffice to explain all the traffic in a ``toy''  sample of 4138 flows. The pairs of packet size and inter-packet interval distributions are almost immediately recognizable as resembling intuitive modes of network behavior. The first basic behavior, for example has all packets near the \caps{MTU}, while intervals are smoothly distributed around a fraction of a millisecond. This is what we would expect from downloading large files; and indeed, the breakdown of packets by port number bears out this interpretation---almost all of the traffic is \caps{HTTP} server traffic, with small slivers of \caps{IMAP} and \caps{SSH} server traffic. The second basic behavior is equally intuitive: tiny packets with frequency distributed around 0.1 seconds---this is all \caps{SSH} traffic, interacting with a remote terminal via keystrokes and screen updates.

@@

The benefits of a performance-preserving algebraic representation are multifold. The algebraic forms of the simplifying assumptions made by many common modeling approaches provide unprecedented clarity and coherence to a complex and confusing subject. For example, assumptions that various aspects of flow behavior are stochastically or deterministically uniform all take the same form algebraically: left matrix multiplication. The other major class of common simplifying assumptions made by traffic models is to apply some marginal behavior across the entire collection of flows or to disjoint subsets of the flows. Such simplifications correspond to right matrix multiplication in \caps{GMM}. Because of the simplicity and clarity that the algebraic structure brings to the subject, we can see clearly now that both common types of simplifying assumptions are na\"ive and simplistic. Our experimental results bolster this conclusion: both uniform and marginal models significantly misrepresent many important network performance metrics. With the general matrix model, however, we stand poised to use the powerful tools of modern linear algebra to develop general, elegant and powerful models of network behavior that just work.

>>>>

% The model of flow behavior implicit in nonnegative matrix factoriation technique is that flow behavior is a composite of some limited set of basic behaviors---and that fractions of each flow ``belong'' to each behavior. This is not an \latin{a priori} unreasonable hypothesis. Contrast that with the assumptions underlying uniform and marginal simplifications of traffic behavior, which can be seen to be intuitively unreasonable without ever conducting and actual experiment or calculation.

\keywords{
  traffic modeling \and
  traffic analysis \and
  workload generation \and
  wireless networks \and
  wireless simulation \and
  linear algebra \and
  nonnegative matrix factorization
}

@@

Once traffic is represented linearly, alternatives to assuming uniformity or marginality become apparent.
We argue that matrix factorization, which is nondestructive and does not assume uniformity or marginality, is far superior to multiplication for this problem.
To demonstrate this approach, we use nonnegative matrix factorization (\caps{NMF})~\cite{Lee01} to find a low-dimensional approximation of the packet size and inter-packet interval matrices of a real traffic trace.
This factorization extracts a small set of ``basic behaviors'' from the traffic matrix and simultaneously explains each flow's observed behavior as a mixture of these basic behaviors.
Although the factorization is computed without examining port numbers, basic behaviors nevertheless correspond to intuitive expectations for protocols.
We find basic behaviors corresponding to typical network activities, including file transfer, typing via \caps{SSH}, web surfing, ping traffic, peer-to-peer and others.

@@

Thus far, we have addressed two out of three of the originally posed questions regarding traffic modeling. First, we have proposed a general approach for representing whole-network traffic without making assumptions of uniformity or marginality. To this end, we introduced the concept of a linear representation of traffic. The general matrix model entails a specific linear representation, such that network traffic patterns can easily be expressed as matrices. Our experimental results show that the \caps{GMM} is sufficiently realistic for generating wireless workloads derived from traces. When uniform or marginal assumptions are applied, however, sufficient realism eludes us, indicating that if we are to simplify and reduce the representation of network behaviors, we must do so by some other technique. In \Section{modeling-concepts}, we observed that application of uniformity and marginality assumptions to traffic instances is equivalent, in the \caps{GMM}, to right and left multiplication, respectively, by matrices of rank one. These transformations destroy almost all of the information in the original traffic matrix.
Conditional models are equivalent to left-multiplication by somewhat less degenerate matrices, but they achieve realism only by approaching the \caps{GMM} itself.

@@

It remains to show how we can simplify and reduce the representation of network behaviors without making unwarranted and manifestly false assumptions of uniformity of marginality of network behaviors. We present here preliminary results from applying nonnegative matrix factorization (\caps{NMF}) to the packet behaviors of real flows. Recall from \Section{general-matrix-model}, that the matrices $\mat{Z}$ and $\mat{V}$ represent the packet size and inter-packet interval histograms of a collection of flows, with each row corresponding to a flow, and each column corresponding to a packet size or a range of inter-packet intervals. We apply \caps{NMF} to the concatenation of these matrices, $\mat{P}=[\mat{Z}~\mat{V}] \in \R^{f \times (d_z+d_v)}$. \caps{NMF} attempts to approximate $\mat{P}$ as the product of three nonnegative matrices:
\begin{align}
\mat{P} \approx \mat{W}\mat{\Sigma}\mat{H},
\end{align}
where $\mat{W} \in \R^{f \times b}$ and $\mat{H} \in \R^{b \times (d_z+d_v)}$ are both row-stochastic, while
$\mat{\Sigma} \in \R^{b \times b}$ is diagonal with descending diagonal entries.
$b\in\N$ is the rank of this factorization, which must be determined somehow as well.
What is the interpretation of such a factorization?
The rows of $\mat{H}$ are \emph{basic behaviors}---pairs of packet size and inter-packet interval distributions, such that each flow's packet behavior can be approximated as a linear combination of these basic behaviors.
Specifically, the matrix $\mat{M}\mat{\Sigma}$ gives the mixing coefficients for each flow.
The diagonal entries of $\mat{\Sigma}$ have a special meaning as well:
they are the number of packets associated with each basic behavior.
Thus the basic behaviors are sorted in descending order of how much traffic they explain.
% Note that the factorization does not associate basic behaviors with individual packets;
% rather, total flow behavior is explained as a mixture of basic behaviors, so the fraction of packets in each flow explained by a given behavior is known, while the specific packets are unknown.

The striking results of applying nonnegative matrix factorization to $\mat{P}$ are shown in \Figure{basic-behaviors}. Twelve basic behaviors suffice to explain all the traffic in a ``toy''  sample of 4138 flows. The pairs of packet size and inter-packet interval distributions are almost immediately recognizable as resembling intuitive modes of network behavior. The first basic behavior, for example has all packets near the \caps{MTU}, while intervals are smoothly distributed around a fraction of a millisecond. This is what we would expect from downloading large files; and indeed, the breakdown of packets by port number bears out this interpretation---almost all of the traffic is \caps{HTTP} server traffic, with small slivers of \caps{IMAP} and \caps{SSH} server traffic. The second basic behavior is equally intuitive: tiny packets with frequency distributed around 0.1 seconds---this is all \caps{SSH} traffic, as indicated by the associate pie chart. This is precisely the behavior we expect from interacting with a remote terminal via keystrokes and screen updates.

% TODO: say something here...

@@

The benefits of a performance-preserving algebraic representation are multifold. The algebraic forms of the simplifying assumptions made by many common modeling approaches provide unprecedented clarity and coherence to a complex and confusing subject. For example, assumptions that various aspects of flow behavior are stochastically or deterministically uniform all take the same form algebraically: left matrix multiplication. The other major class of common simplifying assumptions made by traffic models is to apply some marginal behavior across the entire collection of flows or to disjoint subsets of the flows. Such simplifications correspond to right matrix multiplication in \caps{GMM}. Because of the simplicity and clarity that the algebraic structure brings to the subject, we can see clearly now that both common types of simplifying assumptions are na\"ive and simplistic. Our experimental results bolster this conclusion: both uniform and marginal models significantly misrepresent many important network performance metrics. With the general matrix model, however, we stand poised to use the powerful tools of modern linear algebra to develop general, elegant and effective models of network behavior.
